Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-09-10 analysis/paper/paper.qmd
Updated manuscript

#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
  dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")

@@

metadata.csv <- 
  metadata.csv %>% 
  arrange(site_id)

@@

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g. values below LOD, rounded zeros). Following log-transformation the data was mean-centred and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].

@@

## Near Infrared Spectroscopy (NIRS) {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos.jpg){#fig-nir-pca width=100%, height=100%}

@@

![PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos.jpg){#fig-raman-pca width=100%, height=100%}

@@

![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos.jpg){#fig-xrf-pca width=100%, height=100%}

>>>>

#Filter the sample data to exclude reference samples
tot_samples <- 
  metadata.csv %>% 
  dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartzite")

@@

metadata.csv <- 
  metadata.csv %>% 
  dplyr::filter( type == "Point" | type == "Point fragment" | type == "Preform", 
                 material == "Brecciated quartz" | material == "Quartzite") %>% 
  arrange(site_id)

@@

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g. values below LOD, rounded zeros). Following log-transformation the data was mean-centred and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9; @lundblad_etal08].

@@

## Near Infrared Spectroscopy (NIRS) {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

![NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating three of the light samples](../figures/004-nir-pca-with-photos_v1.jpg){#fig-nir-pca width=100%, height=100%}

@@

![PCA score plot for Raman data. Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot. In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster](../figures/007-raman-pca-with-photos_v1.jpg){#fig-raman-pca width=100%, height=100%}

@@

![XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster based on Fe and Zr content. This results in most of the darker materials clustering towards the bottom of the plot. b) Score plot with > 90% Si samples excluded. Similar to plot A there is a separation based on Fe and Zr values, with darker material towards the bottom of the plot. Light-to-colourless samples remain largely together in cluster 3. Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg](../figures/009-xrf-pca-with-photos_v1.jpg){#fig-xrf-pca width=100%, height=100%}
